---
title: "DATA 624 Project data test"
author: "Samuel I Kigamba"
date: "June 23, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(fpp2)
library(readxl)
```

# This analysis covers "S02Var03" "S03Var05" & "S03Var07"

```{r}
xlsx_path <- "raw_data.xlsx"
df <- readxl::read_xlsx(xlsx_path)
head(df)

```

```{r}
raw = df[c("SeriesInd", "S02Var03", "S03Var05", "S03Var07")]
summary(raw)

```

S03VAr05 and S03VAr07 looks almost identical from the summary above.

Visualize the data and look for seasonality or trend within the different data sets of data.

```{r}

# Data under S02var03 has an outlier which we eliminate before creating visualizations.
raw %>%
  filter(S02Var03 < 20) %>%
  ggplot(aes(x = SeriesInd, y = S02Var03)) +
    geom_line() +
  ggtitle("S02Var03")

raw %>%
  ggplot(aes(x = SeriesInd, y = S03Var05)) +
    geom_line() +
  ggtitle("S03Var05")

raw %>%
  ggplot(aes(x = SeriesInd, y = S03Var07)) +
    geom_line() +
  ggtitle("S03Var07")

```





```{r}
# Convert to time series
SeriesInd.ts <- ts(raw$SeriesInd)
X.ts <- ts(raw$S02Var03)
Y.ts <- ts(raw$S03Var05)
Z.ts <- ts(raw$S03Var07)

```

# S02Var03

```{r}
X1 <- tsCV(raw$S02Var03, ses, h=140)
X2 <- tsCV(raw$S02Var03, holt, h=140)
X3 <- tsCV(raw$S02Var03, holt, damped=TRUE, h=140)

print("S02Var03")
print("MSE")

# Compare MSE:

print(c("ses:", mean(X1^2, na.rm=TRUE)))

print(c("Holt:", mean(X2^2, na.rm=TRUE)))

print(c("Holt damped:", mean(X3^2, na.rm=TRUE)))


print("MAE")
# Compare MAE:

print(c("ses:", mean(abs(X1), na.rm=TRUE)))

print(c("Holt:", mean(abs(X2), na.rm=TRUE)))

print(c("Holt damped:", mean(abs(X3), na.rm=TRUE)))

```

```{r}

X.ts <- raw %>%
  filter(S02Var03 < 20) %>%
  select(S02Var03) %>%
  ts()

X_rwf <- rwf(X.ts, h = 140)
X_drwf <- rwf(X.ts, h = 140, drift = TRUE)
X_mean <- meanf(X.ts, h = 140)

# Add exponential Smoothing and see if its provides better forecasting from the visulization
X_ses <- ses(X.ts, h = 140)

autoplot(X.ts) +
  autolayer(X_rwf, series = "Naive", PI = FALSE) +
  autolayer(X_drwf, series = "Drift", PI = FALSE) +
  autolayer(X_mean, series = "Mean", PI = FALSE) +
  autolayer(X_ses, series = "Ses", PI = FALSE)

```


# Exponential Smoothing

```{r}

X_ses <- ses(X.ts, h = 140)
X_ses[["model"]]

```

This only offers a minor increment in forecasting which is not not justifiable in light of its complexity compared to simple random walk forecast method.


```{r}

# Estimate parameters (Random Walk Forecast)

summary(X_rwf)

```



# S03Var05

```{r}

Y1 <- tsCV(raw$S03Var05, ses, h=140)
Y2 <- tsCV(raw$S03Var05, holt, h=140)
Y3 <- tsCV(raw$S03Var05, holt, damped=TRUE, h=140)


print("S03Var05")
print("MSE")

# Compare MSE:

print(c("ses:", mean(Y1^2, na.rm=TRUE)))

print(c("Holt:", mean(Y2^2, na.rm=TRUE)))

print(c("Holt damped:", mean(Y3^2, na.rm=TRUE)))


print("MAE")

# Compare MAE:

print(c("ses:", mean(abs(Y1), na.rm=TRUE)))

print(c("Holt:", mean(abs(Y2), na.rm=TRUE)))

print(c("Holt damped:", mean(abs(Y3), na.rm=TRUE)))

```


```{r}

Y_rwf <- rwf(Y.ts, h = 140)
Y_drwf <- rwf(Y.ts, h = 140, drift = TRUE)
Y_mean <- meanf(Y.ts, h = 140)

# Add exponential Smoothing and see if its provides better forecasting from the visulization
Y_ses <- ses(Y.ts, h = 140)

autoplot(Y.ts) +
  autolayer(Y_rwf, series = "Naive", PI = FALSE) +
  autolayer(Y_drwf, series = "Drift", PI = FALSE) +
  autolayer(Y_mean, series = "Mean", PI = FALSE) +
  autolayer(Y_ses, series = "Ses", PI = FALSE)

```



# Exponential Smoothing

```{r}

Y_ses <- ses(Y.ts, h = 140)
Y_ses[["model"]]

```

Exponential smoothing only offers a minor increment in forecasting which is not not justifiable in light of its complexity compared to simple random walk forecast method.


```{r}

# Estimate parameters (Random Walk Forecast)

summary(Y_rwf)

```






# S03Var07

```{r}

Z1 <- tsCV(raw$S03Var07, ses, h=140)
Z2 <- tsCV(raw$S03Var07, holt, h=140)
Z3 <- tsCV(raw$S03Var07, holt, damped=TRUE, h=140)


print("S03Var07")
print("MSE")

# Compare MSE:

print(c("ses:", mean(Z1^2, na.rm=TRUE)))

print(c("Holt:", mean(Z2^2, na.rm=TRUE)))

print(c("Holt damped:", mean(Z3^2, na.rm=TRUE)))


print("MAE")
# Compare MAE:

print(c("ses:", mean(abs(Z1), na.rm=TRUE)))

print(c("Holt:", mean(abs(Z2), na.rm=TRUE)))

print(c("Holt damped:", mean(abs(Z3), na.rm=TRUE)))


```



```{r}

Z_rwf <- rwf(Z.ts, h = 140)
Z_drwf <- rwf(Z.ts, h = 140, drift = TRUE)
Z_mean <- meanf(Z.ts, h = 140)

# Add exponential Smoothing and see if its provides better forecasting from the visulization
Z_ses <- ses(Y.ts, h = 140)

autoplot(Z.ts) +
  autolayer(Z_rwf, series = "Naive", PI = FALSE) +
  autolayer(Z_drwf, series = "Drift", PI = FALSE) +
  autolayer(Z_mean, series = "Mean", PI = FALSE) +
  autolayer(Z_ses, series = "Ses", PI = FALSE)

```





# Exponential Smoothing

```{r}

Z_ses <- ses(Z.ts, h = 140)
Z_ses[["model"]]

```

This only offers a minor increment in forecasting which is not not justifiable in light of its complexity compared to simple random walk forecast method.


```{r}

# Estimate parameters (Random Walk Forecast)


summary(Z_rwf)

```